---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs |> summarize(mean_danceability = mean(danceability), sd_danceability = sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_songs |> 
  ggplot() +
    geom_histogram(aes(x = danceability), binwidth = 0.1) 
```

### Task 5: Add a line to the histogram for the mean danceability score

Write code to add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs |> 
  ggplot() +
    geom_histogram(aes(x = danceability), binwidth = 0.1) +
    geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) 
```

### Task 6: Describe our results

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and distribution tell us about the songs in the dataset? 

By looking at the graph creating using this dataset, we can see that though the data closely surrounds the mean, the distribution is skewed slightly to the left. This means the majority of these songs were danceable, as opposed to not, though there are a few outliers at the opposite end of the graf. My biggest question with this data is was there a specific range of years or a specific time period the songs were pulled from? If so, I would write a story maybe describing trends in more uplifting, danceable songs during that era and why those songs might have been prominent. If not, my overarching question woul just fall back on why and how were these specific songs chosen? For this dataset, I think the answer to that question would have a huge hand in my story angle. 

### Task 7: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

